---
title: "snapcount vignette"
author: 
  - name: Christopher Wilks
    affiliation:
    - &jhucs Johns Hopkins University Computer Science
    email: chris.wilks@jhu.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('snapcount')`"
vignette: >
  %\VignetteIndexEntry{snapcount quick start guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  R
---

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()
```

# Basics

## Install snapcount

`R` is an open-source statistical environment which can be easily modified to enhance its functionality via packages. 
`R` can be installed on any operating system from [CRAN](https://cran.r-project.org/) after which you can install snapcount 
by using the following commands in your `R` session:

```{r 'installDer', eval = FALSE}
#Snapcount is not yet in BioConfuctor, 
#you can install it from the GitHub repository directly using devtools
devtools::install_github("langmead-lab/snapr")

```

## Required knowledge

snapcount is based on other packages and in particular in those that have implemented the infrastructure needed for dealing with RNA-seq data. That is, packages like `r Biocpkg('GenomicFeatures')` and `r Biocpkg('recount')` that allow you to import the data. A snapcount user is not expected to deal with those packages directly but will need to be familiar with `r Biocpkg('SummarizedExperiment')` to understand the results snapcount generates.  This vignette is based partially on the [recount2 quick start guide](https://bioconductor.org/packages/release/bioc/vignettes/recount/inst/doc/recount-quickstart.html).

If you are asking yourself the question "Where do I start using Bioconductor?" you might be interested in [this blog post](http://lcolladotor.github.io/2014/10/16/startbioc/).

## Asking for help

As package developers, we try to explain clearly how to use our packages and in which order to use the functions. But `R` and `Bioconductor` have a steep learning curve so it is critical to learn where to ask for help. The blog post quoted above mentions some but we would like to highlight the [Bioconductor support site](https://support.bioconductor.org/) as the main resource for getting help: remember to use the `snapcount` tag and check [the older posts](https://support.bioconductor.org/t/snapcount/). Other alternatives are available such as creating GitHub issues and tweeting. However, please note that if you want to receive help you should adhere to the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). It is particularly critical that you provide a small reproducible example and your session information so package developers can track down the source of the error.

snapcount is an interface to the Snaptron webservices and data.
For questions regarding Snaptron itself and/or its data, please use the [Gitter channel](https://gitter.im/snaptron/Lobby).

## Citing snapcount

We hope that snapcount will be useful for your research. Please use the following information to cite the package specifically, thank you!

```
Christopher Wilks, Rone Charles, Ben Langmead (2019). 
snapcount: R/Bioconductor Package for interfacing with Snaptron 
for rapid quering of expression counts.
https://github.com/langmead-lab/snapr
```

The Snaptron project itself is published separately and can be cited with the following:
```
Wilks, C, Gaddipati, P, Nellore, A, Langmead, B (2018). 
Snaptron: querying splicing patterns across tens 
of thousands of RNA-seq samples. 
Bioinformatics, 34, 1:114-116.
```

# Quick start to using snapcount

Similar to the `r Biocpkg('recount')` all coverage counts in Snaptron/snapcount are stored/retrieved as raw, un-normalized counts.


# Introduction

snapcount is an R package that provides an interface to the [Snaptron web services](http://snaptron.cs.jhu.edu/). 
This package allows you to query and download coverage counts for genes, exons, and exon-exon splice junctions from all (or a subset) of RNA-seq studies.
These RNA-seq studies are organized into compilations (e.g. srav2) that include the same sets of studies that the `r Biocpkg('recount')`  organizes.

The RNA-seq reads from these studies was analyzed through the [Rail-RNA](http://rail.bio) aligner producing the coverage counts used by both `r Biocpkg('recount')` and snapcount.

snapcount complements `r Biocpkg('recount')` in that it allows searching by a specific gene/exon/splice-junction's coordinates or by a HUGO gene across all studies in a specific compilation (e.g. all SRA Human RNA-seq studies in `r Biocpkg('recount')`).

In short, snapcount is gene/region centric whereas `r Biocpkg('recount')` is study centric.

If getting the counts for *all* the genes, exons, and/or splice junctions in an annotation from a particular study is your goal, then we suggest you use the `r Biocpkg('recount')` package instead of snapcount.  

Alternatively, if you're interested in the counts from one or a few genes, but across all (or most) studies/samples in recount2/Snaptron, then you should try snapcount.

This vignette will walk you through examples using the two main areas of the snapcount  package:

* Basic queries to get raw counts from genes, exons, and splice junctions
* High level queries which utilize the basic queries but then do additional summarization (e.g. percent-spliced-in [PSI])


# Basic queries

Basic queries return a *RangedSummarizedExperiment* (RSE) with one or more mapped elements (genes, exons, or splice junctions) as *rowRanges*.
Their raw coverage counts are returned as the *counts* assay in the RSE object.  Full sample metadata is returned as the  *colData* of the RSE object.

The following lists the types of basic queries supported.

* exon-exon splice junctions *query_jx*
* exons *query_exon*
* genes *query_gene*

The Gencode v25 annotation defines what genes and exons can be queried (this is the same as `r Biocpkg('recount')`).
For splice junctions, both annotated and novel junctions can be queried.

Basic queries are appropriate for exploratory analysis or for simply getting the raw counts for a specific set of genes/exons/junctions.

Metadata columns will vary by compilation (e.g. TCGA vs. GTEx).

Also, metadata within a compilation can be incomplete or formatted inconsistently (e.g. "age" in SRAv2).

```{r 'basic', eval = FALSE}
## Load library
library('snapcount')

##Query all exon-exon splice junctions within the region of gene CD99 
#from GTEx v6 sample compilation
#CD99 is chosen to be an example for its manageable size, 
#no specific functional elements are being highlighted.
cd99.jx.all <- query_jx(compilation='gtex', genes_or_intervals='CD99')
head(cd99.jx.all)

##Query coverage for gene, exon, and annotated junctions across all 
#GTEx samples that are labeled with tissue type "Brain" 
#in the region of the CD99 gene

cd99.gene <- query_gene(compilation='gtex', 
                        genes_or_intervals='CD99', 
                        sample_filters='SMTS:Brain')
dim(cd99.gene)
head(cd99.gene)

cd99.exon <- query_exon(compilation='gtex', 
                        genes_or_intervals='CD99', 
                        sample_filters='SMTS:Brain')
dim(cd99.exon)
head(cd99.exon)
###Only query junctions which are fully annotated---both left and 
#right splice sites are found together in one or more of the 
#Snaptron sourced annotations
cd99.jx <- query_jx(compilation='gtex', 
                    genes_or_intervals='CD99', 
                    range_filters='annotated:1', 
                    sample_filters='SMTS:Brain')
dim(cd99.jx)
head(cd99.jx)

##Full compilation metadata (in this case for GTEx) 
#per sample is stored directly in the RangedSummarizedExperient 
#objects returned for each query above.
#For example the library insert size can be retrieved 
#across all runs  in the RSE
head(cd99.jx.all$InsertSize)
```


# High level queries

snapcount can be used to summarize the raw results of the basic queries 
into a more relevant form for downstream analyses via its high level queries.

High level queries include the following:

* Percent spliced in (PSI)
* Junction Inclusion Ratio (JIR)
* Shared sample count (SSC)
* Tissue specificity

The results of high level queries are data frames unless otherwise specified.

## Percent spliced in (PSI)

PSI is a relative measure of usage of a particular splicing variant.

snapcount's version of PSI is currently limited to the common case 
of cassette exons which have two inclusion junctions and one exclusion junction.

The original PSI's values range from 0 to 1 where 0 represents no inclusion coverage for that sample.

Additionally the value of -1 is used in snapcount's PSI function to represent 
samples where either one or the other inclusion groups had 0 coverage or the total raw coverage 
across groups was < min_count (defaults to 20).

```{r 'psi', eval = FALSE}
#Build new query against GTEx
#left inclusion query
lq<-SnaptronQueryBuilder$new(compilation="gtex", 
                             genes_or_intervals="chr19:45297955-45298142", 
                             range_filters="strand=+", 
                             coordinate_modifier=Coordinates$Exact)
#right inclusion query
rq<-SnaptronQueryBuilder$new(compilation="gtex", 
                             genes_or_intervals="chr19:45298223-45299810", 
                             range_filters="strand=+", 
                             coordinate_modifier=Coordinates$Exact)
#exclusion query
ex<-SnaptronQueryBuilder$new(compilation="gtex", 
                             genes_or_intervals="chr19:45297955-45299810", 
                             range_filters="strand=+", 
                             coordinate_modifier=Coordinates$Exact)

psi<-percent_spliced_in(list(lq), list(rq), list(ex))
#order by psi descending
psi<-psi[order(-psi),]
head(psi)
```


## Junction Inclusion Ratio (JIR)

The JIR is a measurement of relative usage similar to the PSI.

However, it allows for more general patterns of splicing to be specified than the Snapcount version of the PSI.

It simply requires two groups of junctions to be defined, these can be overlapping (can include junctions that appear in both groups).

Its values range from -1.0 to 1.0 where 0 represents no particular specificity for either group for that sample.
JIR values approaching 1 represent 0 coverage for groupA and approaching -1 represents 0 coverage for groupB.  
Results are sorted by the JIR statistic (descending).

```{r 'jir', eval = FALSE}
#groupA
A<-SnaptronQueryBuilder$new(compilation="srav2", 
                            genes_or_intervals="chr2:29446395-30142858", 
                            range_filters="strand=-", 
                            coordinate_modifier=Coordinates$Within)
#groupB
B<-SnaptronQueryBuilder$new(compilation="srav2", 
                            genes_or_intervals="chr2:29416789-29446394", 
                            range_filters="strand=-", 
                            coordinate_modifier=Coordinates$Within)

jir<-junction_inclusion_ratio(list(A),
                              list(B),
                              group_names=c("groupA","groupB"))
head(jir)
```

## Shared Sample Count (SSC)

The SSC high level function is designed to support the requirement of many analyses to incorporate multiple splice junctions that are related to the same alternative splicing variant (e.g. an alternatively spliced in cassette exon).  Shared sample count is therefore used to determine if the junctions related to this variant, resulting from the two basic queries the user defines, occur in the same sample.  

If one or more junctions in each query's results occur in a specific sample then that sample is considered "shared", otherwise it's not counted.

The output of the SSC function is a count of samples with this shared property, for each group of two basic junction queries that the user defines.

```{r 'ssc', eval = FALSE}
## We define the left/right splice junction supports of 3 
#cassette exons for the following 3 genes:

## "left"/"right" is relative to the forward strand, 
#reference genome coordinates, 
#we use this terminology instead of 5' or 3' 
#since the gene may be on the reverse strand.

sqb<-SnaptronQueryBuilder
###GNB1
GNB1l<-sqb$new(compilation="gtex", 
               genes_or_intervals="chr1:1879786-1879786", 
               range_filters="strand:-", 
               coordinate_modifier=Coordinates$EndIsExactOrWithin)

GNB1r<-sqb$new(compilation="gtex", 
               genes_or_intervals="chr1:1879903-1879903", 
               range_filters="strand:-", 
               coordinate_modifier=Coordinates$StartIsExactOrWithin)

###PIK3CD
PIK3CDl<-sqb$new(compilation="gtex", 
                 genes_or_intervals="chr1:9664595-9664595", 
                 range_filters="strand:+", 
                 coordinate_modifier=Coordinates$EndIsExactOrWithin)

PIK3CDr<-sqb$new(compilation="gtex", 
                 genes_or_intervals="chr1:9664759-9664759", 
                 range_filters="strand:+", 
                 coordinate_modifier=Coordinates$StartIsExactOrWithin)

###TAP2
TAP2l<-sqb$new(compilation="gtex", 
               genes_or_intervals="chr6:32831148-32831148", 
               range_filters="strand:-", 
               coordinate_modifier=Coordinates$EndIsExactOrWithin)

TAP2r<-sqb$new(compilation="gtex", 
               genes_or_intervals="chr6:32831182-32831182", 
               range_filters="strand:-", 
               coordinate_modifier=Coordinates$StartIsExactOrWithin)

ssc_func<-shared_sample_counts
ssc<-ssc_func(list(GNB1l, GNB1r), 
              list(PIK3CDl, PIK3CDr), 
              list(TAP2l, TAP2r), 
              group_names=c("validated","not validated","validated"))
ssc

```


## Tissue Specificity (TS)

The tissue specificity high level function supports the ability to look for enrichment of tissues across two sets of splicing patterns. 
GTEx is the primary compilation targeted here, but it could work in  other compilations which have complete tissue metadata (possibly TCGA).

The main output of the TS function is a list of all the samples in the compilation with either a "1" in the *shared* column or a "0".
The "1" indicates that there was at least one junction from both groups which occurred in that sample, i.e. the sample is shared across the results of the two junction queries.

The output of this function can be passed to an F-test to measure the significance of the specificity (if present) which uses the number of samples for a tissue
that have the "shared" property set to "1".

A use case of the TS function is to measure the significance of tissue enrichment across junctions which support a cassette exon.

```{r 'ts', eval = FALSE}
groupA<-sqb$new(compilation="gtex", 
                genes_or_intervals="chr4:20763023-20763023", 
                range_filters="strand=-", 
                coordinate_modifier=Coordinates$EndIsExactOrWithin)

groupB<-sqb$new(compilation="gtex", 
                genes_or_intervals="chr4:20763098-20763098", 
                range_filters="strand=-", 
                coordinate_modifier=Coordinates$StartIsExactOrWithin)


ts<-tissue_specificity(list(groupA, groupB))
head(ts)
```




